\thetitle

### 2013-04-13 tfpie/tfpie.tex
Spelling errors.

\begin{abstract}
We propose a bridge between functional and object-oriented programming
in the first-year cirriculum.
\end{abstract}

@@

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
their first year.  Past approaches to achieve this aim have included
teaching an industrial language, such as Java, starting on the first
day.  Others have taken a more indirect route by teaching functional
programming in the first semester, followed by a second semester that
starts in Java.  The latter approach is an improvement over the first,
but both suffer serious flaws.

As an example, Northeastern University teaches functional programming
in the first semester using \emph{How to Design
  Programs}~\cite{dvanhorn:Felleisen2001How}, followed by
object-oriented programming using \emph{How to Design
  Classes}~\cite{local:htdc} in the second semester.  This sequence
was designed to provide a smooth path for incoming students assumed
only to have a competence in high-school level algebra to reach
proficiency in Java by the end of their first year
\cite{dvanhorn:Felleisen2004Structure}.  It was a major improvement
over the previous Java-first cirriculum in terms of student sucess,
attrition, and preparation for subsequent
courses~\cite{dvanhorn:Proulx2006Design}.  However significant
problems remain; in particular, the second semester course violates
the designers' own principles (as recalled
in~\cite{dvanhorn:Bloch2000Scheme}):
%
\begin{enumerate}
\item \emph{introduce only those language constructs that are necessary to
  teach programming principles}, and
\item \emph{choose a language with as
  few language constructs as possible, and one in which they can be
  introduced one at a time}.
\end{enumerate}

@@

\item from a language with a simple structural-equality-based testing
  framework to a language with an industrial strength testing library,

@@

This abrupt transition has several negative consequences: the
principles of object-oriented programming are obscured and
de-emphasized, struggling with the programming environment is
frustrating and can cause potentially good students to leave the
program, it favors students with prior exposure to the particular
tools, it inhibits students from experimenting by relying upon their
past skills, and it creates the false impression that courses are
discrete units of instruction that can be discarded after successful
completion rather than being part of a continuous and culmative
educational experience.

@@

\item Java, even pared down pedagoical Java, is syntactically heavy; discourages
  interactive programming

@@

\begin{itemize}
\item Radically different syntax
\item Types
\item Total committment to objects
\item Subtly different base values
\end{itemize}

@@

Perspective into the upper-class cirriculum:

@@

Matthias Felleisen's approach to pedagody and passion for
undergraduate teaching has inspired this work from the beginning.
CCIS Dean Larry Finkelstein allowed two postdocs to redesign a key
undergraduate course.  Our teaching assistants, Dan Brown, Asumu
Takikawa, and Nicholas Labich, as well as the tutors and graders,
contributed enormously to the success of our courses. Finally, and
most importantly, our students at Northeastern for the last three
years have put up with a curriculum in progress, and the opportunity
to teach them has been truly rewarding.

>>>>

\begin{abstract}
We propose a bridge between functional and object-oriented programming
in the first-year curriculum.
\end{abstract}

@@

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
the students' first year.  Past approaches to achieve this aim have
included teaching an industrial language, such as Java, starting on
the first day.  Others have taken a more indirect route by teaching
functional programming in the first semester, followed by a second
semester that starts in Java.  The latter approach is an improvement
over the first, but both suffer serious flaws.

As an example, Northeastern University teaches functional programming
in the first semester using \emph{How to Design
  Programs}~\cite{dvanhorn:Felleisen2001How}, followed by
object-oriented programming using \emph{How to Design
  Classes}~\cite{local:htdc} in the second semester.  This sequence
was designed to provide a smooth path for incoming students assumed
only to have a competence in high-school level algebra to reach
proficiency in Java by the end of their first year
\cite{dvanhorn:Felleisen2004Structure}.  It was a major improvement
over the previous Java-first curriculum in terms of student success,
attrition, and preparation for subsequent
courses~\cite{dvanhorn:Proulx2006Design}.  However significant
problems remain; in particular, the second semester course violates
the designers' own principles (as recalled
in~\cite{dvanhorn:Bloch2000Scheme}):
%
\begin{enumerate}
\item \emph{introduce only those language constructs that are necessary to
  teach programming principles}, and
\item \emph{choose a language with as
  few language constructs as possible, and one in which they can be
  introduced one at a time}.
\end{enumerate}

@@

@@

This abrupt transition has several negative consequences: the
principles of object-oriented programming are obscured and
de-emphasized, struggling with the programming environment is
frustrating and can cause potentially good students to leave the
program, it favors students with prior exposure to the particular
tools, it inhibits students from experimenting by relying upon their
past skills, and it creates the false impression that courses are
discrete units of instruction that can be discarded after successful
completion rather than being part of a continuous and cumulative
educational experience.

@@

\item Java, even pared down pedagogical Java, is syntactically heavy; discourages
  interactive programming

@@

\begin{itemize}
\item Radically different syntax
\item Types
\item Total commitment to objects
\item Subtly different base values
\end{itemize}

@@

Perspective into the upper-class curriculum:

@@

Matthias Felleisen's approach to pedagogy and passion for
undergraduate teaching has inspired this work from the beginning.
CCIS Dean Larry Finkelstein entrusted two postdocs with the redesign
of a key undergraduate course, which made this experiment possible.
Our teaching assistants, Dan Brown, Asumu Takikawa, and Nicholas
Labich, as well as the tutors and graders, contributed enormously to
the success of our courses. Finally, and most importantly, our
students at Northeastern for the last three years have put up with a
curriculum in progress, and the opportunity to teach them has been
truly rewarding.
